A Bidirected String Graph Model for Genome Assembly

### 2012-12-12 presentation/paper.tex
paper: circulation, Eulerian path

\newcommand{\KeyTerm}[1]{{\it #1}}
\newcommand{\ProgramName}[1]{{\tt #1}}
\newcommand{\BidirectedEdgeForward}[2]{
		\tikz[>=triangle 45,baseline]
		\draw[>->,thin] (0,0.1) node[anchor=east] {#1} --
					    (0.8,0.1) node[anchor=west] {#2};}

@@

However, there are multiple problems with the Eulerian path approach.  First,
the graph may not be Eulerian.  This could happen if the graph is not balanced.
This implies we really would need to solve the bidirected equivalent of the
Chinese Postman Problem instead.\footnote{The Chinese Postman problem on a
bidirected graph is formalized in \cite{Medvedev2009}.}  Even then, the graph
may not be fully connected, which would require that the final walk take the
form of multiple discontinuous walks.  An equally severe problem is that some
overlaps may be spurious, meaning that the two overlapping reads do not actually
form a valid genomic sequence when combined.  Spurious overlaps can be caused by
genomic repeats.  Therefore, we need to allow for the possibility that an edge
in the bidirected graph can be omitted from the walk if it seems like a good
idea to do so.

The way these problems can be solved is to (1) place a lower bound and an upper
bound on the number of times each edge may be traversed, then (2) formulate and
solve a minimum-cost circulation problem that finds the minimum total traversal
count of edges in the graph in order for every vertex to be balanced (i.e. have
the same in-degree as out-degree), then (3) find a generalized Eulerian tour
through the bidirected graph, traversing each edge the exact number of times
that was calculated in (2) and also allowing for the tour be discontinuous.

@@

We would like to use hints from the original data to restrict the number of
times each bidirected edge may be traversed.  This can be done using a
calculation called the \KeyTerm{A-statistic}, or \KeyTerm{arrival-rate
statistic}, which is the logarithm of the probability that an edge represents
unique sequence rather than duplicate sequence \cite{Myers2005}.  The A-statistic
$A(e)$ of an edge $e$ in a bidirected string graph is:

@@

However, $G$, the estimated genome length, is actually unknown, so it must be
estimated first.  It can be estimated by producing a ``bootstrap'' genome length
by adding up the lengths of the sequences on every edge in both directions and
dividing by 2, then calculating the A-statistics using this bootstrap genome
length and identifying edges that are probably single-copy.  The A-statistics
can then be re-estimated over just these single-copy edges in order to refine
the genome length further before calculating the final A-statistic on all the
edges.  This procedure was outlined in \cite{Myers2005}.

@@

A contained read may easily be mapped to the vertex corresponding to a read that
contains it, but this is not sufficient because the A-statistic is calculated
over the edges, not the vertices.  Instead, we need to map each end of each the
contained read to the edge, or in general the set of edges, that contains it.

@@

@@

The A-statistic calculation on each edge $e$, as described in
\Subsection{subsec:a_statistics}, allows us to assign a lower bound $l(e)$ and
an upper bound $u(e)$ on the traversal count for that edge.  Note that $u(e)$
may be $\infty$, as these bounds are very approximate.  We also have $b(v) = 0$
for all $v$, because the flow at every vertex must be balanced; this corresponds
to the walk entering each vertex the same number of times that it is exited,
making the graph as a whole be balanced.

@@

There are still two serious problems with this circulation formulation that need
to be addressed.  First, we actually have a bidirected graph, not a directed
graph.\footnote{The fact that this circulation problem actually needs to be
formulated and solved on a bidirected graph is addressed in \cite{Medvedev2009},
but not in \cite{Myers2005}, even though the latter introduced the overall
fragment string graph algorithm.}

but is a.  We have a
bidirected graph, not a directed graph, so it cannot be trivially mapped to a
directed network flow problem.  A possible workaround for this is to solve
directed flow on the directed equivalent to the bidirected graph, described in
\Subsection{subsec:directed}.  The calculated traversal counts for a directed
edge $e$ between, for example, two vertices $f.E \to g.E$ is not guaranteed to
be the same as the calculated traversal count for the edge that corresponds to
the other direction of the implied bidirected edge, which would be $g.B \to f.B$
in this example.  However, we can take the average of the two counts and use
this as the traversal count for the corresponding bidirected edge, and this

%vertices $f.X$ and $g.X

\subsection{Generalized Eulerian path}

@@

The implementation of the fragment string graph algorithm that was done for this
project is a set of C++ programs that iteratively transform the data on-disk
into the final assembly.  Each program takes as input one or more files as well
as zero or more other command-line arguments and produces as output one or more
files.  Each file contains data such as a list of reads, a list of overlaps, or
a graph.  Internally to the assembler pipeline, data is stored in binary form.
\Figure{fig:impl_graph} shows how the data flows through the main stages of the
assembler.

The input to the assembler is one or more files in plain-text form that each
contain a list of reads.  The final output of the assembler will eventually be a
file in plain-text format that gives the list of assembled contigs.  Currently,
the final stages of the algorithm (\Section{sec:final}, except for the
calculation of the A-statistics) have not been implemented as of the end of the
semester of Fall 2012.

@@

\begin{itemize}
\item Finish the network flow and generalized Eulerian path parts of the
algorithm so that the implementation is ``complete'' and can produce a ``final''
assembly.
\item Improve the performance of \ProgramName{compute-overlaps}.
\item Allow for sequencing errors by finding inexact overlaps and modifying other
stages of the algorithm, such as transitive reduction, to allow for the
possibility that overlaps are inexact.
\item Compensate for sequencing errors by attempting to correct them, either as
part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
\item Consider more recent publications that explore variations on this
algorithm, such as parallelizing the algorithm \cite{Jackson2008}, or using the
Burrows-Wheeler transform to efficiently construct the transitively-reduced
bidirected string graph directly from the reads\cite{Simpson2010}.
\item Use mate-pair information to help resolve repeats and connect assembled
sequences even when they are separated by unknown sequence, possibly using ideas
from \cite{Medvedev2009}.
\end{itemize}

>>>>

\newcommand{\KeyTerm}[1]{{\it #1}}
\newcommand{\ProgramName}[1]{{\tt #1}}
\newcommand{\BidirectedEdgeForward}[2]{
		\tikz[>=triangle 45,baseline]
		\draw[>->,thin] (0,0.1) node[anchor=east] {#1} --
					    (0.8,0.1) node[anchor=west] {#2};}
\newcommand{\BidirectedEdgeInnie}{
		\tikz[>=triangle 45,baseline]
		\draw[>-<,thin] (0,0.1) -- (0.8,0.1) node {\ \ } ;}
\newcommand{\BidirectedEdgeOutie}{
		\tikz[>=triangle 45,baseline]
		\draw[<->,thin] (0,0.1) --(0.8,0.1) node {\ \ };}

@@

However, there are multiple problems with this Eulerian path approach.  First,
the graph may not be Eulerian.  This could happen if the graph is not balanced.
This implies we really would need to solve the bidirected equivalent of the
Chinese Postman Problem instead.\footnote{The Chinese Postman problem on a
bidirected graph is formalized in \cite{Medvedev2009}.}  Even then, the graph
may not be fully connected, which would require that the final walk take the
form of multiple discontinuous walks.  An equally severe problem is that some
overlaps may be spurious, meaning that the two overlapping reads do not actually
form a valid genomic sequence when combined.  Spurious overlaps can be caused by
genomic repeats.  Therefore, we need to allow for the possibility that an edge
in the bidirected graph can be omitted from the walk if it seems like a good
idea to do so.

The way these problems can be solved is to (1) place a lower bound and an upper
bound on the number of times each edge may be traversed, then (2) formulate and
solve a minimum-cost circulation problem that finds the minimum total traversal
count of edges in the graph in order for every vertex to be balanced (i.e. have
the same in-degree as out-degree), then (3) find a generalized Eulerian cycle
through the bidirected graph, traversing each edge the exact number of times
that was calculated in (2), but also allowing for the cycle to be logically
discontinuous by introducing a special vertex and special edges.

@@

We would like to use statistical properties of the original data to restrict the
number of times each bidirected edge may be traversed.  This can be done using a
calculation called the \KeyTerm{A-statistic}, or \KeyTerm{arrival-rate
statistic}, which is the logarithm of the probability that an edge represents
unique sequence rather than duplicate sequence \cite{Myers2005}.  The
A-statistic $A(e)$ of an edge $e$ in a bidirected string graph is:

@@

$G$, the estimated genome length, is actually unknown, so it must be estimated
first.  It can be estimated by producing a ``bootstrap'' genome length by adding
up the lengths of the sequences on every edge in both directions and dividing by
2, then calculating the A-statistics using this bootstrap genome length and
identifying edges that are probably single-copy.  The A-statistics can then be
re-estimated over just these single-copy edges in order to refine the genome
length further before calculating the final A-statistic on all the edges.  This
procedure was outlined in \cite{Myers2005}.

@@

A contained read may easily be mapped to the vertex corresponding to a read that
contains it, but this is not sufficient because the A-statistic is calculated
over the edges, not the vertices.  Instead, we need to map each end of each the
contained read to the edge, or in general the set of edges, that may contain it.

@@

\label{subsec:circulation}

@@

The A-statistic calculation on each edge $e$, as described in
\Subsection{subsec:a_statistics}, allows us to assign a lower bound $l(e)$ and
an upper bound $u(e)$ on the traversal count for that edge.  Note that $u(e)$
may be $\infty$.  We also have $b(v) = 0$ for all $v$, because the flow at every
vertex must be balanced; this corresponds to the walk entering each vertex the
same number of times that it is exited, making the graph as a whole be balanced.
Furthermore, we have $v(e) = 1$ because no edges are considered to be more
costly to use than any other in this formulation.\footnote{Note that the goal of
this assembly algorithm is not to minimize the assembled genome length, but
rather to maximize the similarity of our assembled genome to the true genome.
Therefore, there is no need to assign greater costs to edges that represent
longer DNA sequences (at least in this formulation of the program), because that
would make it more costly to repeat a longer edge even if there is strong
evidence that edge corresponds to repeated sequence.}

@@

There are still two serious problems with this circulation formulation that need
to be addressed.  First, we actually have a bidirected graph, not a directed
graph, so we need to calculate a \KeyTerm{bidirected flow}.\footnote{The fact
that this circulation problem actually needs to be formulated and solved on a
bidirected graph is addressed in \cite{Medvedev2009}, but not in
\cite{Myers2005}, even though the latter introduced the overall fragment string
graph algorithm.}
A bidirected flow is not the same as a directed flow because of the
presence of \BidirectedEdgeInnie and \BidirectedEdgeOutie edges, where the
magnitude of the flow on that edge is counted as going into both vertices (in
the case of a \BidirectedEdgeOutie edge) or out of both vertices (in the case of
an \BidirectedEdgeInnie edge).  This does not imply that the edge itself is a
source or a sink, but rather than the flow is defined in a peculiar way that
makes sense on bidirected graphs.

Fortunately, there does exist a way to solve a bidirected flow problem by
mapping it to a directed flow problem by using the directed equivalent of the
bidirected graph, described in \Subsection{subsec:directed}.  However, the
calculated flow on directed edge $e$ between, for example, two vertices $f.E \to
g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
that corresponds to the other direction of the implied bidirected edge.  This
problem can be worked around by taking the average of the two counts and using
this as the flow on the bidirected edge.  This approximation is equivalent to
performing monotonization on the binet matrix that represents the linear program
reduction of bidirected flow to produce a totally unimodular matrix that
corresponds to a linear program that can be solved to find an integal flow that
is at worst a 2-approximation to the optimal integral
flow\cite{Medvedev2009}\cite{Hochbaum2004}.

The second problem with our circulation formulation is that the bidirected
string graph is not guaranteed to be connected, let alone strongly connected.
For example, there might be vertices that have in-degree 0 but one or more out
edges with nonzero lower bound on flow.  Such vertices may represent the last
read in a genomic region before a coverage gap or the end of the sequence.  Our
goal is to have such vertices be the endpoints of walks through the bidirected
graph.  To take this into account in the formulation of the bidirected flow
problem, we can introduce a special vertex $s$ and unlabeled special edges
between $s$ and every other vertex.  Each special edge $e$ can have any amount
of flow ($l(e) = 0,\ u(e) = \infty$), but at great cost ($v(e) = X$, where $X$
is a very large number).  The addition of these edges will make the graph
strongly connected, and therefore it will be possible to solve the minimum-cost
circulation problem to find a flow that corresponds to balancing the in-degree
and out-degree of every vertex when a certain traversal count $t(e)$ is assigned
to each edge $e$.

\subsection{Generalized Eulerian cycle}

\label{subsec:eulerian}

Here, a \KeyTerm{generalized Eulerian cycle} means a cycle in a bidirected graph
$G$ that traverses each edge $e$ exactly $t(e)$ times, where $t(e)$ is the flow
on edge $e$ as calculated using the method \Subsection{subsec:circulation}.  A
generalized Eulerian cycle is guaranteed to exist because the graph was made
strongly connected before solving the minimum-cost circulation problem.  (A more
thorough explanation of this observation can be found in \cite{Medvedev2009}).
The cycle corresponds to a consistent assembly of the genome, and it can be
decomposed into a set of contigs by placing a contig break every time the
special vertex $s$ is passed through.

It should be possible to find this generalized Eulerian cycle by adapting an
existing Eulerian path algorithm to work on a bidirected graph.

@@

The implementation of the fragment string graph algorithm that was done for this
project is a set of C++ programs that iteratively transform the data on-disk
into the final assembly.  Each program takes as input one or more files as well
as zero or more other command-line arguments and produces as output one or more
files.  Each file contains data such as a list of reads, a list of overlaps, or
a graph.  Internally to the assembler pipeline, data is stored in binary form.
\Figure{fig:impl_graph} shows how the data flows through the main stages of the
assembler.  The assembly pipeline is run using a Makefile that describes this
flow.

The input to the assembler is one or more files in plain-text form that each
contain a list of reads.  The final output of the assembler will eventually be a
file in plain-text format that gives the list of assembled contigs.  However,
the final stages of the algorithm (Sections \ref{subsec:circulation} and
\ref{subsec:eulerian}) have not been implemented as of the end of the semester
of Fall 2012.

@@

\begin{itemize}
\item Finish the minimum-cost circulation and generalized Eulerian path parts of
the algorithm so that the implementation is ``complete'' and can produce a
``final'' assembly.
\item Improve the performance of \ProgramName{compute-overlaps}.
\item Allow for sequencing errors by finding inexact overlaps and modifying other
stages of the algorithm, such as transitive reduction, to allow for the
possibility that overlaps are inexact.
\item Compensate for sequencing errors by attempting to correct them, either as
part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
\item Consider more recent publications that explore variations on this
algorithm, such as parallelizing the algorithm \cite{Jackson2008}, or using the
Burrows-Wheeler transform to efficiently construct the transitively-reduced
bidirected string graph directly from the reads \cite{Simpson2010}.
\item Use mate-pair information to help resolve repeats and connect assembled
sequences even when they are separated by unknown sequence, possibly using ideas
from \cite{Medvedev2009}.
\end{itemize}
